# Results

```{r}
#| echo: false
#| message: false
#| warning: false
library(tidyverse)
library(arrow)
library(scales)
library(RColorBrewer)
library(dplyr)
library(patchwork)
# library(ggridges)
library(vcd)
# library(ggmosaic)
library(lubridate)
library(timeDate)
library(ggrepel)

# Use like: scale_y_continuous(labels = millions) + 
millions <- function(x) {
  ifelse(x == 0, "0", label_comma(scale = 1e-6, suffix = "M")(x))
}

# Use like: scale_y_continuous(labels = thousands) + 
thousands <- function(x) {
  ifelse(x == 0, "0", label_comma(scale = 1e-3, suffix = "K")(x))
}

# Use in mutate: mutate(foo= standardize(bar))
standardize <- function(x) {
  (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)
}

base_color <- brewer.pal(n = 3, name = "Set2")[1]
secondary_color <- brewer.pal(n=3, name = "Set2")[2]
classic <- theme_set(theme_minimal())

temp_label <- "Temperature (°C)"

df <- read_parquet("data/complete_weather_and_taxi_data.parquet")
```

```{r}
# Add a month_abbr column.
df <- df |>
  mutate(month_abbr = factor(month.abb[month], levels = month.abb)) 
```


{{< include sections/3/1_dealing_with_the_pandemic.qmd >}}

```{r}
# Filter out pre- and during-Pandemic data.
df <- df |>
  filter(df$date >= as.Date("2021-10-01"))
```
  
## Patterns in Taxi Data 


{{< include sections/3/2_taxi_usage_patterns.qmd >}}

<!-- ## (3.3) Patterns in weather data? - Moacir -->
## Patterns in Weather Data

{{< include sections/3/3_patterns_in_weather_data.qmd >}}

## Weather vs. Trips.

{{< include sections/3/4_weather_vs_trips.qmd >}}


This is establishing the relationship. Standardize:

( x - min ) / (max - min)

weather: rain y/n , cloud cover categorical , temp numeric

trips: count, cat distance chunks

two variable fights:
  - scatterplot
  - boxplot/cat vs num
  
{{< include sections/3/4_weather_vs_trips_count.qmd >}}
  
## Timeshifts

### last week vs. yesterday

Chat piece about doing this weird choice.

### comparing yesterday to today and lastweek to today.

prob for d3. start with r charts though to see wht seems salient.

---

## Patterns across Hours, Days, and Seasons

In order to determine variations in taxi usage, we needed to establish a baseline "normal" in order to verify fluctuations in the data. How can we infer an abberration if we don't know what normal is? How does cab usage regularly fluctate across the week according to work-life patterns, and how does it fluctuate by season?


