<!-- ## Patterns in Weather Data -->

We chose to limit our weather analysis to three variables: 
temperature, cloud cover, and rain. 
This provides three kinds of variables (numeric, categorical, boolean) 
while also giving a sense of easy to understand information that might be used
by a traveler electing to walk or not. 
That is, a consideration of whether it is sunny or not is perhaps 
clearer to a person choosing to ride the cab than the dew point.

```{r}
df |> 
  ggplot(aes(x = month_abbr, y = temperature)) +
    geom_boxplot(fill=base_color) +
    labs(
      title = "Temperature Distribution",
      x = "Month",
      y = "Temperature (°C)",
    ) 
```
 
 **discussion**
 
 Cloud cover is measured in “oktas,” which we have converted to an ordinal categorical variable, that turns “less cloudy” into a positive measure when comparing two different weather reports.
 
 Value | Description
 ---|---
 0 | Clear
 -1 | Few clouds
 -2 | Scattered clouds
 -3 | Broken clouds
 -4 | Overcast
 
```{r}
#| echo: false
#| message: false
#| warning: false

# Factorize cloud_cover

df <- df |>
  mutate(cloud_cover = factor(cloud_cover, 
    levels = c(0, -1, -2, -3, -4),
    labels = c("Clear", "Few Clouds", "Scattered Clouds", "Broken Clouds", "Overcast"),
    ordered = TRUE))
```
 
 
```{r}
df |>
  group_by(month_abbr, cloud_cover) |>
  tally() |>
  group_by(month_abbr) |>
  mutate(percentage = n / sum(n) * 100) |>
  ungroup() |>
ggplot(aes(x = month_abbr, y = percentage, fill = as.factor(cloud_cover))) +
  geom_bar(stat = "identity", position = "stack") +
  labs(
    title = "Percentage of Cloud Cover by Month",
    x = "Month",
    y = "Percentage",
    fill = "Cloud Cover"
  ) +
  scale_fill_brewer(
    palette = "GnBu"
  ) +
  theme_minimal()
```
 

Cloud cover less obviously seasonal than temperature, and in general days are clear. Likely won't tell us much
otoh makes cloudy days jump out.

```{r}
#| echo: false
#| message: false
#| warning: false

# Booleanize rain 

df <- df |>
  mutate(rain = as.factor(
    ifelse(precipitation != 0, T, F)
    # ifelse(precipitation > 0.1 , T, F)
    )
  )
```

```{r}
df |>
  group_by(date = as.Date(date)) |>
  summarize(
    rain_day = any(rain == TRUE),
    across(c(year, month_abbr), ~ first(.))
  ) |>
  group_by(year, month_abbr) |>
  summarize(
    rainy_days = sum(rain_day), .groups = "drop",
  ) |>
  group_by(month_abbr) |>
  summarize(avg_rainy_days = mean(rainy_days)) |>
  ggplot(aes(x = month_abbr, y = avg_rainy_days)) +
  geom_bar(stat = "identity", fill = base_color) +
  labs(
    title = "Average Number of Days with at Least Trace Amounts of Rain",
    x = "Month",
    y = "Average Rainy Days"
  )
```