<!-- ## Patterns in Weather Data -->

We chose to limit our weather analysis to three variables: 
temperature, cloud cover, and rain. 
This provides three kinds of variables (numeric, categorical, boolean) 
while also giving a sense of easy to understand information that might be used
by a traveler electing to walk or not. 
That is, a consideration of whether it is sunny or not is perhaps 
clearer to a person choosing to ride the cab than the dew point.

```{r}
df |> 
  ggplot(aes(x = month_abbr, y = temperature)) +
    geom_boxplot(fill=base_color) +
    labs(
      title = "Temperature Distribution",
      x = "Month",
      y = temp_label
    ) 
```

As noted earlier, the temperature fluctuates as we would expect. What struck us in these boxplots is how the outliers are distributed, particularly in March, April, and May, which can run hot. This is a fantastic opportunity for seasonal analysis in limiting our investigation, but we have already trimmed our data down so much. Preparing for a specific springtime analysis would require reconsidering how we collect and pre-process our data.

Another point of analysis, however, is to note that everything below -10° or above 30° or so seems to be an outlier. If we consider the distribution all temperatures, we can see that those tails are small and long and might affect the analysis we do moving forward.

```{r}
#| message: false
df |>
  ggplot(aes(temperature)) +
  geom_histogram(color="white", fill=base_color) +
  scale_y_continuous(labels=thousands) +
  labs(
    title = "Distribution of Integral Temperatures", x = "Temperature (°C)", y = "Count")
``` 

Moving forward, we will limit our analysis to temperatures between -5° and 30°.

```{r}
df <- df |>
  filter(temperature >= -5) |>
  filter(temperature <= 30)
```


Cloud cover is measured in “oktas,” which we have converted to an ordinal categorical variable, that turns “less cloudy” into a positive measure when comparing two different weather reports.
 
 Value | Description
 ---|---
 0 | Clear
 -1 | Few clouds
 -2 | Scattered clouds
 -3 | Broken clouds
 -4 | Overcast
 
```{r}
#| echo: false
#| message: false
#| warning: false

# Factorize cloud_cover

df <- df |>
  mutate(cloud_cover = factor(cloud_cover, 
    levels = c(0, -1, -2, -3, -4),
    labels = c("Clear", "Few Clouds", "Scattered Clouds", "Broken Clouds", "Overcast"),
    ordered = TRUE))
```
 
 
```{r}
df |>
  group_by(month_abbr, cloud_cover) |>
  tally() |>
  group_by(month_abbr) |>
  mutate(percentage = n / sum(n) * 100) |>
  ungroup() |>
ggplot(aes(x = month_abbr, y = percentage, fill = as.factor(cloud_cover))) +
  geom_bar(stat = "identity", position = "stack") +
  labs(
    title = "Percentage of Cloud Cover by Month",
    x = "Month",
    y = "Percentage",
    fill = "Cloud Cover"
  ) +
  scale_fill_brewer(
    palette = "GnBu"
  ) +
  theme_minimal()
```
 

Cloud cover less obviously seasonal than temperature, and in general days are clear. The lack of clear seasonality helps us generalize over the year, and the trend toward clear days may help cloudiness jump out more as an excuse to take a cab instead of walk.

```{r}
#| echo: false
#| message: false
#| warning: false

# Booleanize rain 

df <- df |>
  mutate(rain = as.factor(
    ifelse(precipitation != 0, T, F)
    # ifelse(precipitation > 0.1 , T, F)
    )
  )
```

```{r}
df |>
  group_by(date = as.Date(date)) |>
  summarize(
    rain_day = any(rain == TRUE),
    across(c(year, month_abbr), ~ first(.))
  ) |>
  group_by(year, month_abbr) |>
  summarize(
    rainy_days = sum(rain_day), .groups = "drop",
  ) |>
  group_by(month_abbr) |>
  summarize(avg_rainy_days = mean(rainy_days)) |>
  ggplot(aes(x = month_abbr, y = avg_rainy_days)) +
  geom_bar(stat = "identity", fill = base_color) +
  labs(
    title = "Average Number of Days with at Least Trace Amounts of Rain",
    x = "Month",
    y = "Average Rainy Days"
  )
```

In limiting our data just to rush hours, the amount of rain we capture has also been greatly diminished. Where earlier we were averaging about 15 days a month of rain, a lot of that rain fell outside of rush hour, meaning the distribution changes remarkably. This will actually help in later analysis because, like we cloud cover, it makes rain rarer and perhaps a more notable indicator of taxi usage.