<!-- ## Weather vs. Trips. -->

### Influence of Weather on Trip Distances

In addition to limiting our data to trips of ten or fewer miles, we also binned them into groups we thought would be instructive: trips under half a mile, trips up one mile, up to two miles, up to three miles, and up to five miles.

#### Shorter Trips and Higher Temperatures

```{r}
#| message: false
#| warning: false
pivot_trip_distance_df <- df |>
  mutate(temperature = round(temperature)) |>
  mutate(
    up_to_ten_mile_trips = trip_count - rowSums(across(ends_with("mile_trips")))
  ) |>
  pivot_longer(
    cols=c(ends_with("mile_trips")),
    names_to="trip_type",
    values_to="trip_type_count",
  ) |>
  mutate(trip_type = factor(trip_type, 
    levels = c(
      "half_mile_trips", 
      "one_mile_trips", 
      "two_mile_trips", 
      "three_mile_trips", 
      "five_mile_trips", 
      "up_to_ten_mile_trips"
      ),
    labels = c(
      "Up to Half-Mile Trips",
      "Up to One-Mile Trips",
      "Up to Two-Mile Trips",
      "Up to Three-Mile Trips",
      "Up to Five-Mile Trips",
      "Up to Ten-Mile Trips"
    ),
    ordered = TRUE)
  ) 

pivot_trip_distance_df |>
  group_by(trip_type, temperature) |>
  summarize(
    avg_trips_by_degree=mean(trip_type_count),
    .groups = "drop"
  ) |>
  ggplot(aes(temperature, avg_trips_by_degree)) +
    geom_point(color=base_color, size=1, alpha=0.9) +
    scale_y_continuous(
      labels = double_scale,
      breaks = pretty_breaks(n = 3)
    ) +
    geom_smooth(color=secondary_color, method="loess", linewidth=1, se=TRUE, alpha=0.15, span=0.75)+
    labs(
      title = "Average Hourly Trips Per Hour by Temperature across Trip Distance",
      y = "Average Hourly Trips",
      x = temp_label,
      fill = "Count"
    ) +
  facet_wrap(~ trip_type, ncol = 3, nrow = 2, scales="free_y")
```

In retrospect, our distance bins may have been a bit optimistic, 
as no matter how nice it is, there is probably not much of a difference between a five-mile and ten-mile trip if you would rather walk. But on the shorter trips, 
where the choice between taxi and walk might be more salient, especially during rush hour, we can see that sweet spot at around 20° noted in the aggregated data above. The flares in the confidence intervals at the edges indicate the issues of data sparseness, even after we have trimmed by temperature, but for short trips (up to two miles), the sweet spot stands out in the distribution. 

#### It’s Sunny. Let’s Walk.

```{r}
pivot_trip_distance_df |>
  group_by(trip_type, cloud_cover) |> 
  summarize(
    avg_trips = mean(trip_type_count, na.rm = TRUE),
    se = sd(trip_type_count, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  ) |>
  ggplot(aes(x = cloud_cover, y = avg_trips, fill = cloud_cover)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~trip_type, scales = "free_y") +
  scale_y_continuous(labels=double_scale)+
  scale_fill_brewer(palette = "GnBu") +
  labs(
    title = "Average Hourly Trips by Cloud Cover and Trip Distance",
    x = "Cloud Cover",
    y = "Average Hourly Trips",
    fill = "Cloud Cover"
  )+
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    legend.position = "bottom" 
 ) 
```

In comparing average numbers of trips for different cloud covers, the distinctions are subtle, but notable, especially comparing short trips to long. For trips of up to one mile, we see that cloudy circumstances have higher averages than clear days, while the distinction is far more muted in longer trips.

#### On Not Destroying One’s Clothing Walking in the Rain

```{r}
pivot_trip_distance_df |>
  group_by(trip_type, rain) |> 
  summarize(
    avg_trips = mean(trip_type_count, na.rm = TRUE),
    se = sd(trip_type_count, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  ) |>
  ggplot(aes(x = rain, y = avg_trips, fill = rain)) +
  geom_bar(stat = "identity", position = "dodge") +
  # geom_errorbar(aes(ymin = avg_trips - se, ymax = avg_trips + se),
  #               width = 0.2, position = position_dodge(0.9)) +
  scale_y_continuous(labels=comma)+
  scale_fill_manual(values=c("grey", rain_color), labels="Not raining", "Raining") +
  labs(
    title = "Average Hourly Trips by Rain and Trip Distance",
    x = "Rain",
    y = "Average Hourly Trips",
    fill = ""
  )+
  facet_wrap(~trip_type, scales = "free_y") +
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    legend.position = "bottom" 
 ) 
```

This distribution marks perhaps most clearly the relationship that trip distance probably has as a background cause of whether one walks or not. If we assume no one wants to walk in the rain, the distinction between rainy hours and non rainy hours for short trips makes perfect sense. But for trips over five miles, the distinction washes out, suggesting rain plays no role in deciding whether to take a five mile or longer trip.