### Count of trips 


```{r}
#| echo: false
#| message: false
#| warning: false
library(arrow)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(RColorBrewer)
library(scales)
library(patchwork)
library(ggrepel)

base_color <- brewer.pal(n = 3, name = "Set2")[1]
rain_color <- brewer.pal(n = 3, name = "Set2")[3]
classic <- theme_set(theme_minimal())

df <- read_parquet("../../data/complete_weather_and_taxi_data.parquet")
df <- df |> 
  filter(date >= as.Date("2021-10-01"))

# Establishing working day and rush hour
library(timeDate)

all_dates <- seq.Date(as.Date("2021-01-01"), as.Date("2024-12-31"), by = "day")
us_holidays <- as.Date(holidayNYSE(c(2019:2024)))

get_columbus_day <- function(year) {
  october_dates <- seq.Date(as.Date(paste(year, "10-01", sep = "-")), as.Date(paste(year, "10-31", sep = "-")), by = "day")
  mondays <- october_dates[weekdays(october_dates) == "Monday"]
  mondays[2]
}
get_black_friday <- function(year) {
  november_dates <- seq.Date(as.Date(paste(year, "11-01", sep = "-")), as.Date(paste(year, "11-30", sep = "-")), by = "day")
  thursdays <- november_dates[weekdays(november_dates) == "Thursday"]
  thanksgiving <- thursdays[4] 
  thanksgiving + 1 
}
christmaseve_to_new_years <- as.Date(unlist(lapply(2021:2024, function(year) {
  seq.Date(as.Date(paste(year, "12-24", sep = "-")), as.Date(paste(year, "12-31", sep = "-")), by = "day")
})))

columbus_days <- as.Date(sapply(2021:2024, get_columbus_day))
black_fridays <- as.Date(sapply(2021:2024, get_black_friday))

working_days <- all_dates[
  !weekdays(all_dates) %in% c("Saturday", "Sunday") & 
  !(all_dates %in% us_holidays) & 
  !(all_dates %in% columbus_days) & 
  !(all_dates %in% black_fridays) & 
  !(all_dates %in% christmaseve_to_new_years) & 
  (format(as.Date(all_dates), "%m-%d") != "11-11")
]

df$working_day <- ifelse(df$date %in% working_days, T, F)
df$rush_hour <- ifelse(df$working_day==T & df$hour %in% c(7,8,9,17,18,19), T, F)
df <- df |> 
  filter(rush_hour == T) |> 
  mutate(
    rush_hour_group = as.factor(case_when(
      hour %in% c(7, 8, 9) ~ "Morning rush hour (7-9am)",
      hour %in% c(17, 18, 19) ~ "Evening rush hour (5-7pm)"
    )),
    rain = as.factor(ifelse(precipitation != 0, "Raining", "Not raining")),
    cloud_cover.f = factor(
      case_when(
        cloud_cover == 0 ~ "Clear",
        cloud_cover == -1 ~ "Few",
        cloud_cover == -2 ~ "Scattered",
        cloud_cover == -3 ~ "Broken",
        cloud_cover == -4 ~ "Overcast"
      ),
      levels = c("Clear", "Few", "Scattered", "Broken", "Overcast"),
      ordered = TRUE
    )
  )
```


```{r}
ggplot(df, aes(x = temperature, y = trip_count)) +
  geom_point(alpha = 0.2, color = base_color) +
  geom_smooth(method = "loess", se = T, color = base_color) +
  labs(
    title = "Trip Count vs. Temperature",
    x = "Temperature (°C)",
    y = "Trip Count"
  ) +
  theme_minimal()
```
We see a higher amount of variability in the lower temperatures. We see an increase in trip count at higher temps, specifically above 30* celsius. This may make sense. If it is hot out maybe I dont want to walk. 


```{r}
ggplot(df, aes(x = cloud_cover.f, y = trip_count) )+
  geom_point(position="jitter", alpha=0.3, color=base_color) +
  geom_boxplot(alpha=0.1, fill="white", outlier.alpha=1, outlier.color="black", linewidth=0.55) +
  labs(title = "Trip Count by Cloud Cover",
       x = "Cloud Cover",
       y = "Trip Count") +
  theme_minimal() 

```
The vast majority of days are either clear or overcast. This doesn't really tell us much - those values are very similar, the middle plots could really just be explained by variability and smaller sample size. potensh remove unless we just want for due diligence. 



```{r}
#| fig-width: 10
#| fig-height: 6
ggplot(df, aes(x = trip_count, fill = rain)) +
  geom_density(alpha = 0.7) +
  labs(
    title = "Density of trip count by Rain Status",
    x = "Number of trips in an hour",
    y = "Density",
    fill = "Rain Status"
  ) +
  scale_fill_manual(values = c("Not raining" = "grey", "Raining" = "cornflowerblue")) +
  theme_minimal()

```
The density curves for both "Not Raining" and "Raining" largely overlap, suggesting that rain has some, but not a substantial, impact on trip count distribution.
The Raining distribution appears slightly shifted towards higher trip counts compared to Not Raining - this could indicate a slight increase in trip demand during rainy periods, potentially due to fewer people walking or biking. The tail is slightly longer on the right side, suggesting that extreme trip counts (higher demand) might occur more frequently during rain.

Perhaps: 


```{r}
ggplot(df, aes(x = rush_hour_group, y = trip_count, fill = rain)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) + #
  labs(
    title = "Trip Count distribution by rain and rush_hour group",
    x = "",
    y = "Trip Count",
    fill = "Rain Status"
  ) +
  scale_fill_manual(values = c("Not raining" = "grey", "Raining" = "cornflowerblue")) +
  scale_color_manual(values = c("Not raining" = "grey", "Raining" = "cornflowerblue")) +
  theme_minimal()

```

Potentially illuminates that people are slightly more rigid in morning routines compared to evening routines. 



