<!-- ## Timeshifts -->

With these provocative initial conclusions, we decided to move forward to see if we could relativize the concept of “nice.” 
The issue here became so perceptual, however, we are not certain we solved a way out of it. After all, “it’s too nice out to take a cab” implicitly invokes a rather recent moment in time when it was not nice. Perhaps it is too nice since it is abnormally warm, or sunny after a week of clouds. Or sunny in the early evening after a rainy morning. We could track those kinds of patterns. However, mapping it onto the taxi data presents special challenges.

That is, the distinct variability from day-to-day of taxi usage, even limiting ourselves to rush hour on weekdays, means we cannot chain multi-day patterns as in the distinctions above. 

So we invented a workaround and wanted to see how it would help. 
To wit, we compared hourly trip counts on a given day to the same hour on the same day of the week the previous week. 
Additionally, we compared weather data to the same hour of the previous day. Our hope was that this would capture some of the compressed sense of “nice” as opposed to “unnice” with a comparison to a similar day for ridership. Perhaps we would see a similar effect as above.

We also decided to stick to analyzing temperature alone, given time considerations.

```{r}
#| message: false
#| warning: false
df |>
  ggplot(aes(temperature_change_since_prev_day, trip_count_change_since_prev_week, fill=temperature)) +
  geom_point(alpha=.15, size=3, stroke=0, shape=21 ) +
  scale_fill_distiller(palette="GnBu", direction=-1) +
  scale_y_continuous(labels=thousands) +
  labs(
    title = "Change in Hourly Trip Count by Change in Temperature",
    x = paste("Change in ", temp_label, "Since 24 Hours Ago"),
    y = "Change in Trips Since a Week Ago",
    fill = paste("“Current”", temp_label)
  ) 
```

The distribution here suggests that trips don’t change much when the temperature also does not change much. If we squint, we imagine more dots in the II quadrant (negative $x$, positive $y$), which confirms our idea that people ride less when it is nicer (read: warmer) out. But maybe it is our imagination.

```{r}
df |> 
  mutate(
    Quadrant = case_when(
      temperature_change_since_prev_day > 0 & trip_count_change_since_prev_week > 0 ~ "Got Warmer & More Trips (I)",
      temperature_change_since_prev_day > 0 & trip_count_change_since_prev_week < 0 ~ "Got Warmer & Fewer Trips (IV)",
      temperature_change_since_prev_day < 0 & trip_count_change_since_prev_week > 0 ~ "Got Cooler & More Trips (II)",
      temperature_change_since_prev_day < 0 & trip_count_change_since_prev_week < 0 ~ "Got Cooler  & Fewer Trips (III)"
    )
  ) |>
  count(Quadrant) |>
  arrange(desc(n))
```

While we do not see much of an effect here, the effect we do see matches our intuition. In general, warmer weather means fewer taxi rides, while cooler weather occasions more trips. We will try to reason out this relativization in the next section. 

```{r}
#| message: false
#| warning: false
df |>
# pivot_trip_distance_df |>
  # select(month, trip_count, trip_type, trip_type_count, temperature, temperature_change_since_prev_day) |>
  select(month, trip_count, temperature, temperature_change_since_prev_day) |>
  mutate(temperature = round(temperature)) |>
  mutate(temperature_change_since_prev_day = round(temperature_change_since_prev_day)) |>
  write_csv("data-for-d3.csv")
```